---
title: 'STATS 60 Summer 2020: HW1'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Due on Friday July 3rd at 5PM PDT on Gradescope
## 1) Summarizing Data

Consider the following dataset

$$
\{6,8,7,9,3,7,7,10,8,2\}
$$

1) What is the mean?

2) What is the median?

3) What is the mode?

4) What is the standard deviation?

## 2) Introduction to R

1) Assign the character string "10" to `x` and "20" to `y`

2) What happens when you run `x + y`? Why does this happen?

3) How would we treat `x` and `y` as numbers to add them?

4) Suppose I wanted to print the string "STATS60". I run the following code. Why do I get the following error? What should I do instead?

```{r eval = FALSE}
print(STATS60)
```

`## Error in print(STATS60) : object 'STATS60' not found`

Suppose now I wanted to print the string "060". I run the following code. Why do I get the incorrect output? What should I do instead?

```{r}
print(060)
```

## 3) R for Data

1) Create a vector `v` of the data from Part 1.

2) Create a histogram of `v`

3) Create a boxplot of `v`. What is your interpretation of this?

4) Write code that gives you the mean and standard deviation of `v`.